Supernova Host Galaxies in the Dark Energy Survey: \\ II. Rapidly Evolving Transients

### 2020-02-24 RET_Hosts.tex
Updates from Overleaf

% Abstract of the paper
\begin{abstract}
Rapidly evolving transients (RETs) are a mysterious class of astrophysical event. They are characterised by lightcurves that decline much faster than standard supernovae (SNe), span vast ranges in peak luminosity and can be seen to redshifts greater than 1. 
\end{abstract}

@@

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{figs/cum_SFR_mike.png}
\caption{Cumulative distributions of the SFR of RET hosts, compared to CCSNe and the low-z SDSS sample.
\label{fig:sfr_cum}}
\end{figure}

@@

We derive our sample from the 106 RETs discovered in the 5-year DES-SN transient survey. This number expands upon the 72 of \citetalias{Pursiainen2018}. The first reason for the increased sample size is the use of the 5th year of DES-SN, where \citetalias{Pursiainen2018} were only able to make use of the first four years. The second reason is an update to the sample selection technique. In \citetalias{Pursiainen2018}, objects were highlighted based on various lightcurve criteria such as time of rise to peak brightness ($t_{\mathrm{rise}}$), time of decline ($t_{\mathrm{decl}}$), and lightcurve `full-width half-maximum' ($t_{\mathrm{FWHM}}$). Candidates fitting these criteria were then vetted manually in order to make the final sample. The manual vetting included a removing a large number of active galactic nuclei (AGN). In this work, we expand upon this procedure with the introduction of a machine learning technique

We refer to the full sample as DES RETs. Of the 106 objects in the sample, 97 have a host galaxy detected in deep host galaxy photometry of \citet{Wiseman2020}, of which 49 have a host galaxy spectroscopic redshift. A further three have redshifts obtained from narrow lines observed in spectra of the transients themselves. We do not consider these three objects for the analysis, since we are unable to separate transient and host contributions to the spectra.

@@

\subsection{Estimating metallicities \label{subsec:calc_Z}}

@@

\subsection{Metallicity \label{subsec:res_metallicity}}
\begin{figure}
\includegraphics[width=0.5\textwidth]{figs/RET_MZR.png}
\caption{The mass-metallicity relation (MZR) for RET host galaxies.
\label{fig:mzr}}
\end{figure}

@@

@@

In Fig.~\ref{fig:mzr} we show the mass-metallicity relation (MZR) for the RET and comparison samples.

@@

\subsection{Transient - host galaxy correlations}

@@

\subsection{Metallicity}

The sample of DES RETs shows a preference for low metallicity host environments. The PDF displays a strong similarity to the hosts of SNe Ibc, as well as LGRBs. There is a clear difference to the PDFs of SNe II, which follow SDSS field galaxies, while the aversion of high-metallicity systems is not as strong as for SLSNe. This result is suggestive of a stripped-envelope, massive-star origin for RETs.

>>>>

% Abstract of the paper
\begin{abstract}
Rapidly evolving transients (RETs) are a mysterious class of astrophysical event. They are characterised by lightcurves that decline much faster than standard supernovae (SNe), span vast ranges in peak luminosity and can be seen to redshifts greater than 1. Their evolution on fast timescales has hindered high quality follow-up observations, such that their origin and explosion/emission mechanism remains unexplained. In this paper, we investigate the host galaxies of the largest RET sample to date from the Dark Energy Survey (DES). Using deep-stacked photometry and emission-lines from OzDES spectroscopy, we derive host galaxy stellar mass, star-formation rate (SFR), and metallicity for 42 hosts. We find that RETs explode almost exclusively in star-forming galaxies and are thus likely associated with massive stars. Comparing RET hosts to samples of host galaxies of other explosive transient as well as field galaxies, we find that RETs prefer galaxies with high specific SFR, indicating a link to young stellar populations stripped-envelope SNe. RET hosts appear to show a lack of chemical enrichment, their metallicities akin to long duration gamma-ray bursts and superluminous SNe host galaxies.

\end{abstract}

@@

\begin{figure*}
\includegraphics[width=\textwidth]{figs/RET_Mosaic.png}
\caption{Selection of DES RET host galaxies in an RGB composite of the DES $gri$ bands. The locations of the transients are indicated with cyan crosses.
\label{fig:mag_dist}}
\end{figure*}

@@

@@

We derive our sample from the 106 RETs discovered in the 5-year DES-SN transient survey. This number expands upon the 72 of \citetalias{Pursiainen2018}. The first reason for the increased sample size is the use of the 5th year of DES-SN, where \citetalias{Pursiainen2018} were only able to make use of the first four years. The second reason is an update to the sample selection technique. By imposing the \citetalias{Pursiainen2018} selection criteria on season 5, the sample is increased to 96 objects.

\subsection{Improvements to the search method}

The original method of finding RETs in the DES-SN data (and presented in \citetalias{Pursiainen2018}) was designed to be simple and used light curve modelling with Gaussian and linear fits. The simplistic method made it possible to look for exotic transients without knowing their observed characteristics beforehand and resulted in a large sample of photometrically selected fast transients. However, as the search was simplistic and relied heavily on visual inspection of the available data (etc. images, light curves, host galaxy information), it is impossible to be sure how complete the sample is. For instance, due to the large redshift range within the sample it is entirely possible that distant events could have been missed due to time dilation. Here, a more sophisticated search method is presented. As only a fraction of transients in DES-SN have redshift information from host galaxy, the search must be done in the observer frame.  The key features of RETs that distinguishes them from most traditional SNe types are the fast light curve evolution and blue colour at peak. Even though both of these quantities depend on the redshift of the transient, they still seem to distinguish the the fast events from traditional SN types.

To assess the distinguishing photometric properties of RETs, we take the sample of 72 from the \citetalias{Pursiainen2018} method along with spectroscopically confirmed SNe of types Ia, Ibc, II, and SLSNe from DES that pass the following criteria:
\begin{enumerate}
\item The transient was only detected in one DES-SN observing season.
\item Maximum observed brightness in both $g$- and $r$-bands was brighter than 24 mag (in the eight `shallow' DES-SN fields) or 25 mag (in the two `deep' fields).
\item $g$- and $r$-band observations used for the colour had to be taken within 2 days in observer frame.
\end{enumerate}

Of objects passing these cuts, the RETs cluster at shorter timescales and bluer colours than other SNe, even in the observer frame. However, using photometric data points directly has several problems that can be improved. For one, measuring peak colour is problematic: DES-SN did not always observe $g$- and $r$-bands on the same or even consecutive nights, thus making it impossible to measure the peak colour in a number of cases. Measuring rise times of 10-15 days is difficult to perform with a one week cadence when it has to be done without fitting a light curve model, hence the rise time values are spread over a wide range. To negate this issue, we interpolate the light curves using the method of Gaussian processes (GP). The interpolated light curves have a 0.5 day cadence and every epoch has a flux value, and an associated uncertainty, for each band. Using this technique, SNe Ia and RETs populate two distinct regions of $g-r$-$t_{\mathrm{rise; r}}$ parameter space and are thus easier to separate. We define a region in this parameter space which minimises the contamination of non-RETs (purity) while maximising the total fraction of RETs (completeness). The resulting limits are $-1.5 < g-r < 0.6 $ and $t_{\mathrm{rise; r}} <16.5$.

We process all $\sim 30,000$ DES-SN transient candidates with GP. In order to reduce the contamination from active galactic nuclei (AGN), we use a basic convolutional neural network (CNN) classifier which we train on spectrosopically confirmed objects, and use to separate the sample into two subtypes: AGN-like and SNe-like. The classifier returns SNe-like objects with an accuracy of 0.992 on the test set; the remaining AGN are likely removed by the final manual vetting.
The SNe passing the CNN classifier are subjected to the LC quality cuts, resulting in 2259 objects, of which 939 lie inside the colour and rise-time region which we defined as RETs. These objects are subject to a further set of cuts. We impose a cut based on a fit of the LC with the \texttt{PSNID} software (\citet{Sako2008}). We use thresholds of \texttt{FITPROB}$<0.91$ and \texttt{PBAYES}$<0.82$ to remove highly-probably SNe Ia, which removes 46 objects from the RET parameter space. In order to further remove longer-lived SNe, the decline time to half of the peak brightness must be $<24$ days. This removes 347 SNe, resulting in 546 objects remaining inside the parameter space. The final 564 transients have been visually inspected, with the majority rejected for being spurious detections, obvious multi-season variability that was not picked up by the CNN, or showing evidence for a longer timescale decline.

Using the above method recovers N of the 96 rets found using the \citetalias{Pursiainen2018} technique, and adds a further M.
We refer to the resulting sample as DES RETs. Of the 106 objects in the sample, 97 have a host galaxy detected in deep host galaxy photometry of \citet{Wiseman2020}, of which 49 have a host galaxy spectroscopic redshift. A further three have redshifts obtained from narrow lines observed in spectra of the transients themselves. We do not consider these three objects for the analysis, since we are unable to separate transient and host contributions to the spectra.

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{figs/cum_SFR_mike.png}
\caption{Cumulative distributions of the SFR of RET hosts, compared to CCSNe and the low-z SDSS sample.
\label{fig:sfr_cum}}
\end{figure}
\subsection{Estimating metallicities \label{subsec:calc_Z}}

@@

\subsection{Metallicity \label{subsec:res_metallicity}}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{figs/RET_MZR.png}
\caption{The mass-metallicity relation (MZR) for RET host galaxies.
\label{fig:mzr}}
\end{figure}

@@

In Fig.~\ref{fig:mzr} we show the mass-metallicity relation (MZR) for the RET and comparison samples. The contours show the MZR for low-redshift ($\hat{z}=0.08$) star-forming galaixes from SDSS, adjusted to the PP04 O3N2 diagnostic. We use the MZR parameterisation \citet{Zahid2014} to show the best fit to the MZR for star-forming galaxies. The blue dashed line shows the fit to the low-z data, while the orange dashed line corresponds to the MZR at $z=0.45$, the mean redshift of the RET host sample. The RET hosts lie systematically below the galaxy MZR fits as well as the bulk of the SDSS galaxies, meaning that for a given stellar mass they have a lower metallicity.

@@

@@

\subsection{Origin of RETs}

The sample of DES RETs shows a preference for low-metallicity, strongly star-forming host environments. The PDF of their metallcities displays a strong similarity to the hosts of SNe Ibc, as well as LGRBs. There is a clear difference to the PDFs of SNe II, which follow SDSS field galaxies. The preference for low-metallicity systems is not as strong as for LGRBs or SLSNe, but the highest metallicities found in all three samples are very similar at around solar metallicity. This result is further suggestive of a stripped-envelope, massive-star origin for RETs. 
The population of RET hosts lies, on average, between CCSNe and LGRBs/SLSNe in terms of both star formation and metallicity. A loose correlation exists between the luminosity and rarity of events, and the host galaxy conditions required for their formation. The rough rate of RETs ($\geq 10^{-6} \mathrm{Mpc}^{-3} \mathrm{yr}^{-1}$), (\citetalias{Pursiainen2018}) is $\sim1\%$ of the CCSN rate \citep{Li2011,Horiuchi2011}, which itself is divided into the more common SNe II and sub-dominant SESNe \citep{Kelly2012,Frohmaier2020}. At $\sim1\%$ of the CCSN rate, RETs are more common than SLSNe ($\sim0.01 - 0.05\%$ of CCSNe; \citealt{McCrum2015,Prajz2016,Frohamier2020}) and LGRBs (intrinsically $\sim0.08\%$ when accounting for beaming; \citealt{Graham2016}). These figures place the DES RETs between extreme objects (SLSNe, LGRBs) and more common SNe (SNe II, SESNe) in terms of rate, matching their location in host galaxy parameter space. While stressing that these associations are lose - rates are uncertain and host galaxy parameters span wide ranges for all transients - they are both linked to the respective transients' progenitor channels. Based upon both indicators, it is reasonable to infer that RETs are linked to very massive stars that also require some extreme properties such as rapid rotation, albeit not as much as progenitors as SLSNe or LGRBs. It could therefore be possible that RETs are an intermediate and/or precursory step, whereby the initial collapse of the star occurs leading to a shocked photosphere, but conditions are not highly tuned enough for a LGRB or SLSN and the respective central engine does not form.

\subsection{Correlations with lightcurve properties}
